---
title: Clustering and differential expression analysis using Monocle
---

```{r}
library(monocle3)
library(rjson)
source("../../libraries/utils.R")
```

Data loading and preparation

```{r}
dataset_name = 'PBMC1'  # VARIABLE
DATASET_PATH = "../../dataset/"
DATA_DIR = paste(DATASET_PATH, dataset_name, '-Filtered/10X', sep='')
LABEL_DIR = paste(DATASET_PATH, dataset_name, '-Filtered', sep='')
OUT_DEFAULT_DIR = paste("../../results/", dataset_name, '/monocle/default/', sep='')
OUT_CELLTYPIST_DIR = paste("../../results/", dataset_name, '/monocle/celltypist/', sep='')
OUT_PROTEIN_DIR = paste("../../results/", dataset_name, '/monocle/protein/', sep='')
if (!dir.exists(OUT_DEFAULT_DIR)) {
  dir.create(OUT_DEFAULT_DIR, recursive = TRUE)
}
if (!dir.exists(OUT_CELLTYPIST_DIR)) {
  dir.create(OUT_CELLTYPIST_DIR, recursive = TRUE)
}
if (!dir.exists(OUT_PROTEIN_DIR)) {
  dir.create(OUT_PROTEIN_DIR, recursive = TRUE)
}

N_CLUSTERS_CELLTYPIST = fromJSON(file=paste('../../dataset/', dataset_name, '-Filtered/raw/nclusters.json', sep=''))$nclusters
N_CLUSTERS_PROTEIN = fromJSON(file=paste('../../dataset/', dataset_name, '-Filtered/10x/nclusters.json', sep=''))$nclusters

TOP_MARKER_NUM = 500
```

```{r}
cds <- load_mm_data(
  mat_path = paste(DATA_DIR, "matrix.mtx", sep = "/"), 
  feature_anno_path = paste(DATA_DIR, "genes.tsv", sep = "/"),
  cell_anno_path = paste(DATA_DIR, "barcodes.tsv", sep = "/")
)
```

Studying feature variance

```{r}
cds <- preprocess_cds(cds, num_dim = 10, method = "PCA") # normalization
plot_pc_variance_explained(cds)
```

Clustering with default parameters

```{r}
cds <- cluster_cells(cds, reduction_method = "PCA")
cat(paste("Number of clusters:", length(levels(clusters(cds, reduction_method = "PCA")))))
```

Save the default clustering results

```{r}
label_df = data.frame(clusters(cds, reduction_method = "PCA"))
label_df$cell <- rownames(label_df)
rownames(label_df) <- 1:nrow(label_df)
colnames(label_df)[1] <- "computed_id"
label_df$computed_id <- as.numeric(label_df$computed_id)
write_clustering(OUT_DEFAULT_DIR, label_df, "cell", "computed_id", dist(t(exprs(cds))))
```


Differential expression on default clustering results

```{r}
marker_test_res <- top_markers(
  cds,
  group_cells_by="cluster",
  genes_to_test_per_group = 3000,
  reduction_method='PCA',
  cores=10
)
```

Save markers

```{r}
write_markers(OUT_DEFAULT_DIR, marker_test_res, "gene_id", "cell_group", "marker_test_p_value", FALSE, TOP_MARKER_NUM)
```

Clustering tuning resolution according to celltypist

```{r}
cat("Clusters to find:")
N_CLUSTERS_CELLTYPIST
```


```{r}
# get ids of clusters bigger than 40 cells
mapping = read.csv(paste('../../dataset/', dataset_name, '-Filtered/raw/celltypist_mapping.csv', sep=''))
counts = read.csv(paste('../../dataset/', dataset_name, '-Filtered/raw/celltypist_annotation_counts.csv', sep=''))

mapping_counts = merge(mapping, counts, by.x = "go", by.y = "cluster.ids")
mapping_counts = subset(mapping_counts, count > 40)
clusters_ids_to_keep = mapping_counts$id

# get barcodes of cells in clusters bigger than 40 cells
celltypist_labels = read.csv(paste('../../dataset/', dataset_name, '-Filtered/raw/celltypist_labels.csv', sep=''))
celltypist_labels = subset(celltypist_labels, cluster.ids %in% clusters_ids_to_keep)
barcodes_to_keep = celltypist_labels$cell

cds <- load_mm_data(
  mat_path = paste(DATA_DIR, "matrix.mtx", sep = "/"), 
  feature_anno_path = paste(DATA_DIR, "genes.tsv", sep = "/"),
  cell_anno_path = paste(DATA_DIR, "barcodes.tsv", sep = "/")
)
# keep only cells in clusters bigger than 40 cells
cds = cds[, colData(cds)$cell %in% barcodes_to_keep]
```

```{r}
cds <- preprocess_cds(cds, num_dim = 10, method = "PCA") # normalization
plot_pc_variance_explained(cds)
```


```{r}
range_upper = 3
range_lower = 0

repeat{
  resolution = (range_upper + range_lower) / 2
  cat(paste("Trying resolution ", resolution, sep=''))
  cds <- cluster_cells(cds, reduction_method = "PCA", resolution=resolution)
  num_clusters = length(levels(clusters(cds, reduction_method = "PCA")))
  cat(paste("Got ", num_clusters, ' clusters', sep=''))
  if (num_clusters == N_CLUSTERS_CELLTYPIST){
    break
  }
  else if (num_clusters < N_CLUSTERS_CELLTYPIST){
    # increase the resolution
    range_lower = resolution
  }
  else{
    # reduce the resolution
    range_upper = resolution
  }
}
```


Save the tuned clustering results

```{r}
label_df = data.frame(clusters(cds, reduction_method = "PCA"))
label_df$cell <- rownames(label_df)
rownames(label_df) <- 1:nrow(label_df)
colnames(label_df)[1] <- "computed_id"
label_df$computed_id <- as.numeric(label_df$computed_id)
write_clustering(OUT_CELLTYPIST_DIR, label_df, "cell", "computed_id", dist(t(exprs(cds))))
```

Differential expression on tuned clustering results

```{r}
marker_test_res <- top_markers(
  cds,
  group_cells_by="cluster",
  genes_to_test_per_group = 3000,
  reduction_method='PCA',
  cores=10
)
```

Save markers

```{r}
write_markers(OUT_CELLTYPIST_DIR, marker_test_res, "gene_id", "cell_group", "marker_test_p_value", FALSE, TOP_MARKER_NUM)
```

Clustering tuning resolution according to protein surface

```{r}
cat("Clusters to find:")
N_CLUSTERS_PROTEIN
```

```{r}
# get barcodes of cells labelled using protein surface
protein_labels = read.csv(paste('../../dataset/', dataset_name, '-Filtered/10x/labels.csv', sep=''))
barcodes_to_keep = protein_labels$cell

cds <- load_mm_data(
  mat_path = paste(DATA_DIR, "matrix.mtx", sep = "/"), 
  feature_anno_path = paste(DATA_DIR, "genes.tsv", sep = "/"),
  cell_anno_path = paste(DATA_DIR, "barcodes.tsv", sep = "/")
)
cds = cds[, colData(cds)$cell %in% barcodes_to_keep]
```

```{r}
cds <- preprocess_cds(cds, num_dim = 10, method = "PCA") # normalization
plot_pc_variance_explained(cds)
```


```{r}
range_upper = 3
range_lower = 0

repeat{
  resolution = (range_upper + range_lower) / 2
  cds <- cluster_cells(cds, reduction_method = "PCA", resolution=resolution)
  num_clusters = length(levels(clusters(cds, reduction_method = "PCA")))

  if (num_clusters == N_CLUSTERS_PROTEIN){
    break
  }
  else if (num_clusters < N_CLUSTERS_PROTEIN){
    # increase the resolution
    range_lower = resolution
  }
  else{
    range_upper = resolution
  }
}
```

Save the tuned clustering results

```{r}
label_df = data.frame(clusters(cds, reduction_method = "PCA"))
label_df$cell <- rownames(label_df)
rownames(label_df) <- 1:nrow(label_df)
colnames(label_df)[1] <- "computed_id"
label_df$computed_id <- as.numeric(label_df$computed_id)
write_clustering(OUT_PROTEIN_DIR, label_df, "cell", "computed_id", dist(t(exprs(cds))))
```

Differential expression on tuned clustering results

```{r}
marker_test_res <- top_markers(
  cds,
  group_cells_by="cluster",
  genes_to_test_per_group = 3000,
  reduction_method='PCA',
  cores=10
)
```

Save markers

```{r}
write_markers(OUT_PROTEIN_DIR, marker_test_res, "gene_id", "cell_group", "marker_test_p_value", FALSE, TOP_MARKER_NUM)
```

```{r}
sessionInfo()
```
